Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-02-20 writeups/Su_IGW_break.tex
everything but new movie stuff is checked in

% TODO
% - Figure 1: Could you include a tick mark at t=0? Does the curve start at t=0?
% Section 6.2 needs more narrative discussion of the results. We should add another paragraph saying that the typical Richardson number of the critical layer is between 0.25 and 0.5, characteristic of Kelvin-Helmholtz instabilities. This suggestions the critical layer width is self-consistently determined by the ``turbulence’’ in our simulations. The layer width does not change with resolution. We should also say that here we calculate Ri using the background buoyancy frequency, because the fluctuations do not change N^2 significantly; thus, Ri~0.5 can be used to solve for the critical layer width in a star, as we know N^2 and U_c (corresponding to the tidal frequency Omega).
% Investigate a smarter time averaging for the "average over everything
% predictor"?
    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

@@

A full video of our simulation with $N_x = 768$, $N_z = 3072$, $\mathrm{Re} =
1024$ is available
online\footnote{http://www.princeton.edu/~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

@@

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min_z
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max_z
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
To understand the variation in $\mathrm{Ri}$ over $x$, we also compute
$\min\limits_x \mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown
in Fig.~\ref{fig:nl_f_ri}. The critical layer quickly narrows to its minimum
width, driven by the anti-diffusivity of IGWs.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number (Eq.~\eqref{eq:ri_med_def}) of the flow at
    the critical layer over time (in units of $N^{-1}$) in our fiducial
    simulation. The red and solid green lines denote respectively the minimum
    and median of $\mathrm{Ri}_x(x, t)$. These numbers measure the mean and
    spread in width of the critical layer over $x$. Note that $\mathrm{Ri} \sim
    \frac{1}{4}$ corresponds to the KHI, so this plot suggests the shear at the
    critical layer does not steepen past the onset of the
    KHI.}\label{fig:nl_f_ri}
\end{figure}

@@

In Fig.~\ref{fig:nl_front} we plot the numerically measured $z_c$ against two
semi-analytic models: (i) integration of Eq.~\eqref{eq:zc_anal} using the
measured $F_a(t)$, and (ii) substituting the time-averaged $\ev{F_a}_t$ (over
the entire length of the simulation) into Eq.~\eqref{eq:zc_sol}. Since $z_c(t)$
is less well-defined at early times (when the critical layer is thick and
transient behavior is strong), we solve Eq.~\eqref{eq:zc_anal} by integrating
backwards from the end of the simulation ($t = t_f$), using $z_c(t_f)$ as the
initial condition. From Fig.~\ref{fig:nl_front}, we see that the agreement
between the measured $z_c(t)$ and its estimate via $F_a(t)$ and
Eq.~\eqref{eq:zc_anal} is excellent.

By time-averaging the numerically measured $F_a$, we find $\ev{F_a}_t \approx
0.71F_{al}$, which is used for the second model in Fig.~\ref{fig:nl_front}.
Note that $F_a < F_{al}$, so momentum flux absorption at the critical layer is
incomplete. This is due to reflection of waves off the critical layer, which
carry momentum downward.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_front.png}
    \caption{Propagation of the critical layer over time. Shown are: (black)
    $z_c(t)$ from simulation data, (green) model for $z_c(t)$ using direct
    integration of Eq.~\eqref{eq:zc_anal} for $F_a(t)$ measured from simulation
    data (described in Eq.~\eqref{eq:fa_def}), and (blue) direct substitution of
    time-averaged $\ev{F_a(t)}_t$ into Eq.~\eqref{eq:zc_sol}. The models use the
    end of the simulation as initial conditions and integrate backwards, as
    $z_c$ is less well-defined at early times. The agreement of the
    directly-integrated model with the data shows Eq.~\eqref{eq:zc_anal} is a
    good description of the evolution of $z_c$. The the poorer but qualitatively
    correct agreement of the time-averaged model with the data shows both that
    $F_a(t) < F_{al}$.}\label{fig:nl_front}
\end{figure}

>>>>

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

@@

A full video of our simulation with $N_x = 768$, $N_z = 3072$, $\mathrm{Re} =
1024$ is available
online\footnote{http://www.princeton.edu/\~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

@@

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
We use the background buoyancy frequency to compute $\mathrm{Ri}$, as
fluctuations do not change $N^2$ significantly ($\sim 1\%$). To understand the
variation in $\mathrm{Ri}$ over $x$, we also compute $\min\limits_x
\mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown in
Fig.~\ref{fig:nl_f_ri}. The Richardson number at the critical layer quickly
decreases to a typical value between $0.25$ and $0.5$, characteristic of the
onset of the KHI\@.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number (Eq.~\eqref{eq:ri_med_def}) of the flow at
    the critical layer over time (in units of $N^{-1}$) in our fiducial
    simulation. The red and solid green lines denote respectively the minimum
    and median of $\mathrm{Ri}_x(x, t)$. These numbers measure the mean and
    spread in width of the critical layer over $x$. Note that $\mathrm{Ri} \sim
    \frac{1}{4}$ corresponds to the KHI, so this plot suggests the shear at the
    critical layer does not steepen past the onset of the
    KHI.}\label{fig:nl_f_ri}
\end{figure}

The steadiness of $\mathrm{Ri}$ at this characteristic value suggests that the
critical layer width is self-consistently determined by the ``turbulence'' in
our simulations. The width does not change significantly with resolution in our
resolved simulations (see Fig.~\ref{fig:agg}). Thus, in stars, $\mathrm{Ri} \sim
0.5$ can be used to calculate the critical layer width, where $N^2$ and
$\overline{U}_c$ (corresponding to the tidal frequency) are known.

@@

In Fig.~\ref{fig:nl_front} we plot the numerically measured $z_c$ against
numerical integration of Eq.~\eqref{eq:zc_anal} using the measured $F_a(t)$.
Since the critical layer is still forming at early times, we solve
Eq.~\eqref{eq:zc_anal} by integrating backwards from the end of the simulation
($t = t_f$), using $z_c(t_f)$ as the initial condition. From
Fig.~\ref{fig:nl_front}, we see that the agreement between the measured $z_c(t)$
and its estimate via $F_a(t)$ is excellent.

By time-averaging the numerically measured $F_a$, we find $\ev{F_a}_t \approx
0.71F_{al}$. Note that $F_a < F_{al}$, so momentum flux absorption at the
critical layer is incomplete. This is due to reflection of waves off the
critical layer, which carry momentum downward.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_front.png}
    \caption{Propagation of the critical layer over time. Shown are: (black)
    $z_c(t)$ from simulation data, and (red) model for $z_c(t)$ using direct
    integration of Eq.~\eqref{eq:zc_anal} for $F_a(t)$ measured from simulation
    data (described in Eq.~\eqref{eq:fa_def}). The model uses the end of the
    simulation as its initial condition and integrates backwards, as the
    critical layer is still forming at earlier times. The agreement of the model
    with the simulation shows Eq.~\eqref{eq:zc_anal} is a good description of
    the evolution of $z_c$.}\label{fig:nl_front}
\end{figure}
